---
title: "Rana boylii Population Genetics"
output: 
  flexdashboard::flex_dashboard:
    css: flexstyles.css
    social: [ "twitter", "github" ]
    source_code: "https://https://github.com/ryanpeek/ryanpeek.github.io"
    theme: yeti
    logo: ../img/fylfrog_icon_resize.png
---

<!-- date: '<small>Updated: `r format(Sys.time(), "%Y-%b")`</small>' -->

```{r setup, include=FALSE}

suppressPackageStartupMessages({
  library(flexdashboard);
  library(sf);
  library(leaflet);
  library(here);
  library(viridis);
  library(mapview);
  library(plotly)
})
knitr::opts_chunk$set(echo = FALSE, warning=FALSE)
knitr::opts_knit$set(root.dir = normalizePath("../"))
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

Overview {.storyboard data-icon="fa-leaf"}
=========================================

### Freshwater organisms are rapidly declining globally, and over 1/3 of all amphibian taxa are at risk of extinction. *Rana boylii* is a stream breeding frog native to CA and OR.  {data-commentary-width=400}

```{r, out.width = '100%'}
knitr::include_graphics(paste0(here(),"/img/rabo_A_P1060730.jpg"))
```

***

#### **BACKGROUND**

 - *Rana boylii* provide a unique link with hydroclimatic change as their breeding phenology is tightly coupled with river hydrology. They have evolved strong breeding associations with changes in flow associated with rain or snowmelt cues in the spring.
 - As one of the only obligate stream-breeding frogs in western North America, the foothill yellow-legged frog (*Rana boylii* Baird 1854) historically occurred in streams and rivers from Oregon to northern Baja California west of the Sierra-Cascade crest across a range of elevations from sea level to ~1400 m.
 - *Rana boylii* have declined over 50% from the species former range (Davidson et al. 2005, Lind 2005). Many existing populations are small.
 - Small populations with limited genetic diversity may have reduced adaptive potential and difficulty responding to future environmental change. Identifying these populations is crucial for effective conservation prioritization and management.

#### **KEY QUESTIONS**

 1. Can we better characterize range-wide genetic structure of *R. boylii*, and identify areas for prioritization?
 1. Has river (flow) regulation reduced connectivity and genetic health (diversity) of *R. boylii*?
 1. Can we quantify this genetic signature for specific hydrologic flow regimes?
 




### Historical range for *Rana boylii* was broad, extending from Southern California all the way into Oregon, and from the coast to the Sierra Nevada. {data-commentary-width=400}

```{r, out.width = '100%', out.height= '100%', echo=F, warning=F, message=F}

# fig.width=10, fig.height=6.5}
suppressPackageStartupMessages({
  #library(leaflet);
  library(tidyverse);
  library(sf);
  library(mapview)
})  

# Rana boylii range
load(paste0(here(),"/data/RABO_Potential_Range_CAandOR.rda"))

# All RABO Points
load(paste0(here(),"/data/RABO_all_records_thru_2010_v4.rda"))
mapview(rabo_range, col.regions="orange", alpha.regions=0.5) + mapview(rabo_pts, col.regions="yellow")
  
  
```



### Altered flow regimes have affected freshwater species that evolved under natural flow patterns by limiting connectivity, reducing flow cues, and eliminating habitat.


```{r out.width='90%'}
knitr::include_graphics(paste0(here(),"/img/fig_01b_mean_ann_discharge_combined_noRUB_labs_resized.png"))
knitr::include_graphics(paste0(here(),"/img/recession_seasonality_breeding.png"))
```

***

#### Flow Regimes 

 - Different flow regimes exihibit different patterns of magnitude, duration, frequency, and variability. 
 - These components can be quantified in many different ways. 
 - Foothill yellow-legged frogs have cued in on specific periods of the hydrograph to spawn, typically the spring snowmelt/rain recession.


### Flow seasonality & predictability changes with regulation 

```{r, out.width = '90%'}
knitr::include_graphics(paste0(here(),"/img/pre_cowplot_predic_seas_annot.png"))
knitr::include_graphics(paste0(here(),"/img/post_cowplot_predic_seas_annot.png"))

```

***

#### Flow Seasonality and Predictability shifts with flow regulation.

 - Using flow data coupled with wavelet analysis and Colwell's M/P metric of seasonality has been shown to be a useful way to compare the inter and intra-annual changes in flow regimes (Tonkin et al. 2017). 
 - Strong patterns of seasonality (within-year) and predictability (among-years) can be observed in unregulated rivers (Figure A). 
 - Using data from USGS gaging stations, I ran a similar analysis looking at many years of flow data before and after dams were built in the Yuba and American Rivers
 - We can see significant shift in seasonality in rivers that are regulated (Figure B).




### Massive parallel sequencing (genetic) data can be used to evaluate many aspects of at risk populations including population size, genetic diversity, connectivity, migration, etc.

```{r, out.width = '95%'}
knitr::include_graphics(paste0(here(),"/img/rapture_slides_1.jpg"))
```

***

#### RADSeq/Rapture is flexible and cost efficient

 - Able to sequence thousands of samples by multiplexing at once (high throughput)
 - The basic steps are:
    - Digest with Sbf1
    - Add RAD adapters
    - Sonicate/shear
    - Purify
    - Add/select specific capture baits/loci of interest 

#### Using magnetic nanobeads to wash/extract DNA

<iframe width="560" height="315" src="https://www.youtube.com/embed/PfihhGqB7Vc?rel=0" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

Rivers {data-icon="fa-tint"}
=========================================

Column {data-width=800}
-----------------------------------------------------------------------

### River Stage on NFA

```{r nfa}
# load data
load(paste0(here(),"/data/2011-2015_solinst_mainstem_hourly_compensated.rda"))
nfa<- hrly2 %>% 
  filter(site=="NFA" & WY==2015) %>%
  select(Datetime, Level, wyd, Temperature)
(nfa2 <- ggplot() +
    geom_line(data=nfa, aes(x=Datetime, y=Level, color=Temperature)) + theme_bw() +
    scale_color_viridis("Water Temperature (C)",breaks=c(seq(0,30,3)), limits=c(0,30)) + ylab("Stage (m)") +
    theme(legend.spacing = unit(0.1, "in"), legend.direction = "horizontal",
          legend.justification = c("right", "top"),legend.box.just = "right",
          legend.position = c(.95, .95))+
    guides(color = guide_colourbar(title.position ="top", barheight = 0.75, barwidth = 12)))

# nfa3 <- ggplot() +
#     geom_line(data=nfa, aes(x=Datetime, y=Level, color=Temperature)) + theme_bw() +
#     scale_color_viridis("Water Temperature (C)",breaks=c(seq(0,30,3)), limits=c(0,30)) + ylab("Stage (m)")
# ggplotly(nfa3)

```

### River Stage on MFA

```{r mfa}
# load data
load(paste0(here(),"/data/2011-2015_solinst_mainstem_hourly_compensated.rda"))
mfa<- hrly2 %>% 
  filter(site=="MFA" & WY==2015) %>%
  select(Datetime, Level, wyd)
mfa2 <- ggplot() + geom_line(data=mfa, aes(x=Datetime, y=Level), color="gray40") +
  geom_point(data=mfa, aes(x=Datetime, y=Level, color=Level),
             pch=21, fill="gray", alpha=0.7) + theme_bw() + ylab("Stage (m)") + 
  scale_color_viridis()
ggplotly(mfa2)

```

Research {.storyboard data-icon="fa-map"}
=========================================


Photos {data-icon="fa-picture-o"}
=========================================

Column {data-width=700}
-----------------------------------------------------------------------

#### **Rivers have a Story to Tell**

> *"The face of the water, in time, became a wonderful book&mdash;a book that was a dead language to the uneducated passenger, but which told its mind to me without reserve… And it was not a book to be read once and thrown aside, for it had a new story to tell every day”* - Mark Twain, Two Views of the Mississippi, 1883

```{r out.height= '50%', fig.align='center'}
knitr::include_graphics(paste0(here(),"/img/indian_ck.JPG"))
```


Column {data-width=300}
-----------------------------------------------------------------------

###

```{r out.height= '50%'}
knitr::include_graphics(paste0(here(),"/img/IMG_9098.JPG"))
```
