---
title: "Quoted"
categories: blog
date: "`r format(Sys.time(), '%Y-%m-%d')`"
layout: post
bibliography: '../Peek-bibkey-2016.bib'
---

feature and teaser image paths start at website/images/

images in post are specified as "/images/blog/..."

Save in website/_Rmd as post-name.rmd, source _Rutil/knitpost.R and call knitAll or knitPost


### RMarkdown blogging

 - .Rmd files go in `_Rmd`
 - `source(Rutils/knitpost`
 
 - call `knitPost('file-name')` or `knitAll()`
 - .md file should land in _posts/blog/
 - On `bundle exec jekyll build`, .html should end up in `_site/blog/new-post_title/index.html`
 - image paths are specified as images/, so specify, e.g. as "blog/my-image.png"

 
```{r}
source("./_Rutil/knitpost.R")
#knitPost('2017-11-21-mapping-with-sf-part-3.Rmd')
knitPost('2018-11-13-a-few-mapping-tips.Rmd')


```

### Preview Site with Jekyll

Use `bundle exec jekyll serve` to build a local version of the website. You can also use `bundle exec jekyll serve --no-watch` to avoid the auto watch for changes. To end, just use `bundle exec jekyll serve --detach` to kill process.

```{r refs, warning=FALSE}
library(RefManageR, quietly = T, verbose = F)
bib <- ReadBib("../Peek-bibkey-2016.bib") # Change to whatever is your .bib file
#myopts <- BibOptions(bib.style = "authoryear", style="latex", first.inits=FALSE, max.names = 20)
#Citet(bib, author="lind", .opts=list(cite.style="numeric", super=TRUE))

```

## Required Readings

Read this stuff [@Lind2011, @Kupferberg1996]

## Citations
