---
title: "Quoted"
categories: blog
date: "`r format(Sys.time(), '%Y-%m-%d')`"
comments: true
date: "`r format(Sys.time(), '%Y-%m-%d')`"
layout: post
tags: [leaflet, field]
---

```{r setup, include=FALSE}
library(ggplot2)
library(plotly)
library(dygraphs)
library(dplyr)
library(viridis)
```


feature and teaser image paths start at website/images/

images in post are specified as "/images/blog/..."

Save in website/_Rmd as post-name.rmd, source _Rutil/knitpost.R and call knitAll or knitPost


### RMarkdown blogging

#- .Rmd files go in _Rmd
#- source Rutils/knitpost
#- call knitpost('file-name') or knitAll()
#- .md file should land in _posts/blog/
#- On `bundle exec jekyll build`, .html should end up in _site/blog/new-post_title/index.html
#- image paths are specified as images/, so specify, e.g. as blog/my-image.png

```{r echo=F, warning=F, message=F}

library(leaflet); library(rgdal); library(dplyr)

# Projection
WGS84 <- "+proj=longlat +ellps=WGS84 +datum=WGS84"

t1 <- readOGR("data/2016-06-28-NFA-Sailor.GPX", layer="tracks", verbose = FALSE)
p1 <- readOGR("data/2016-06-28-NFA-Sailor_points.GPX", layer="waypoints", verbose = FALSE)

p1<-spTransform(p1, CRS(WGS84))
t1<-spTransform(t1, CRS(WGS84))

mapSailor <- leaflet() %>% addTiles() %>% 
  setView(lng = -120.49312, lat = 39.209, zoom = 12) %>%  # set to Auburn/Colfax, zoom 5 for CA 
  addTiles(group = "OSM") %>%
  addProviderTiles("Stamen.TopOSMFeatures", group = "OSM Features") %>%
  addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") %>%
  addProviderTiles("Thunderforest.Landscape", group = "Topo") %>%
  hideGroup("OSM Features") %>% 

  # PhD samples
  addCircleMarkers(data=p1, group="Rana boylii",
                   popup = paste0("Name: ", 
                                  p1@data$name, "<br>","Description: ",
                                  p1@data$desc),
                   stroke=TRUE, weight=0.6,radius=11,
                   fillOpacity = 0.8, color="black",
                   fillColor = "yellow") %>%
  
  # Sailor Canyon hike
  addPolylines(data=t1, group="Sailor Trail",
               popup = paste0("Name: ", t1@data$name),
               stroke=TRUE, weight=0.9,
               fillOpacity = 0.8, color="maroon") %>%
  
  # add controls for basemaps and data
  addLayersControl(
    baseGroups = c("OSM", "ESRI Aerial", "Topo"),
    overlayGroups = c("Rana boylii",
                      "Sailor Canyon Trail", "OSM Features"),
    options = layersControlOptions(collapsed = T))

mapSailor
  
```

#### Script:
```{r echo=T, eval=FALSE, message=FALSE}
library(leaflet); library(rgdal); library(dplyr)
# Projection  
WGS84 <- "+proj=longlat +ellps=WGS84 +datum=WGS84"  
# read in GPX  
t1 <- readOGR("data/2016-06-28-NFA-Sailor.GPX", layer="tracks")  
p1 <- readOGR("data/2016-06-28-NFA-Sailor_points.GPX", layer="waypoints")  
# transform
p1<-spTransform(p1, CRS(WGS84))
t1<-spTransform(t1, CRS(WGS84))
# make a map!
m <- leaflet() %>% addTiles() #this is shortened
setView(lng = -120.49312, lat = 39.209, zoom = 10) %>%  # set to Auburn/Colfax, zoom 5 for CA 
  addTiles(group = "OSM") %>%
  addProviderTiles("Stamen.TopOSMFeatures", group = "OSM Features") %>%
  addProviderTiles("Esri.WorldImagery", group = "ESRI Aerial") %>%
  addProviderTiles("Thunderforest.Landscape", group = "Topo") %>%
  hideGroup("OSM Features") %>% 
# PhD samples
  addCircleMarkers(data=p1, group="Rana boylii",
                   popup = paste0("Name: ", 
                                  p1@data$name, "<br>","Description: ",
                                  p1@data$desc),
                   stroke=TRUE, weight=0.6,radius=11,
                   fillOpacity = 0.8, color="black",
                   fillColor = "yellow") %>%
  # Sailor Canyon hike
  addPolylines(data=t1, group="Sailor Canyon Trail",
               popup = paste0("Name: ", t1@data$name),
               stroke=TRUE, width=0.9,
               fillOpacity = 0.8, color="maroon") %>%
  # add controls for basemaps and data
  addLayersControl(
    baseGroups = c("OSM", "ESRI Aerial", "Topo"),
    overlayGroups = c("Rana boylii",
                      "Sailor Canyon Trail", "OSM Features"),
    options = layersControlOptions(collapsed = T))
m
```
