---
title: "website_management"
categories: blog
date: "`r format(Sys.time(), '%Y-%m-%d')`"
layout: post
bibliography: '../Peek-bibkey-2016.bib'
---

## Dealing with `ruby`

To keep things up to date, highly recommend using `rbenv`. Once installed, can use this to check for available versions:

    `rbenv install -l`

Then install new version of ruby:
    `rbenv install 2.7.2`

### check ruby version and use rehash to set paths:

check the version: `ruby -v`
rehash: `rbenv rehash`

### set the global version to use

`rbenv global 2.7.2`
`ruby -v`
`rbenv versions`

## Update Jekyll & Bundler

First we need to install bundler and jekyll to build up our site:

 - `gem install bundler jekyll`
 
Now we can actually run stuff in the directory where the `Gemfile` lives:

 - `bundle update`
 - `bundle install`

# Update and View website!

To preview the website, we can use this:

 - `bundle exec jekyll serve`
 
If we have a drafts folder and want to see what the posts in drafts would look like, use this:

 - `bundle exec jekyll serve --drafts`

## Locations

feature and teaser image paths start at website/images/

images in post are specified as "/images/blog/..."

Save in website/_Rmd as post-name.rmd, source _Rutil/knitpost.R and call knitAll or knitPost


### RMarkdown blogging

 - .Rmd files go in `_Rmd`
 - `source('_Rutil/knitpost.R')`
 
 - call `knitPost('file-name')` or `knitAll()`
 - .md file should land in _posts/blog/
 - On `bundle exec jekyll build`, .html should end up in `_site/blog/new-post_title/index.html`
 - image paths are specified as images/, so specify, e.g. as "blog/my-image.png"

 
```{r}
source("./_Rutil/knitpost.R")
#knitPost('2017-11-21-mapping-with-sf-part-3.Rmd')
knitPost('2021-09-04-distill-websites.Rmd')

```

### Preview Site with Jekyll

Use `bundle exec jekyll serve` to build a local version of the website. You can also use `bundle exec jekyll serve --no-watch` to avoid the auto watch for changes. To end, just use `bundle exec jekyll serve --detach` to kill process.

```{r refs, warning=FALSE}
library(RefManageR, quietly = T, verbose = F)
bib <- ReadBib("../Peek-bibkey-2016.bib") # Change to whatever is your .bib file
#myopts <- BibOptions(bib.style = "authoryear", style="latex", first.inits=FALSE, max.names = 20)
#Citet(bib, author="lind", .opts=list(cite.style="numeric", super=TRUE))

```

## Required Readings

Read this stuff [@Lind2011, @Kupferberg1996]

## Citations
